\pkg{evtree}: Evolutionary Learning of Globally Optimal Classification and Regression Trees in \proglang{R}

### 2011-09-20 pkg/evtree/inst/doc/evtree.Rnw
BBBClub example
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@243 255f4249-2263-4417-b1ea-8498779e27a7

This section continues with a sample session where \code{evtree} is used, in conjunction with \code{rpart} and \code{ctree}, to analyse the \textit{Bookbinder's Book Club} data. In this session, all trees are constrained to have a minimum number of $26$ observations per terminal node ($2\%$ of the total number of observations). To avoid excessively large trees \code{ctree} is called with $\alpha = 1\%$. To provide uniform visualizations and predictions of the fitted models the \code{as.party()} function is used to convert \pkg{rpart} objects into objects of class \pkg{party}. The sample session starts with the construction and visualization of trees using the \code{rpart} and \code{ctree} methods.
%
<<rpart-ctree>>=
data("BBBClub", package = "evtree")
library("rpart")
set.seed(1090)
rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 26))
ct <- ctree(choice ~ ., data = BBBClub, mincrit = 0.99, minbucket = 26)
plot(rp)
plot(ct)
@
%
The resulting trees are displayed in Figure~\ref{fig:example_rpart_ctree}.
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\centering
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{rpart} \end{flushleft}
\centering
<<rp-plot, fig=TRUE, echo=FALSE, width=12, height=6.8>>=
plot(rp)
@
\end{minipage}
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{ctree} \end{flushleft}
\centering
<<ct-plot, fig=TRUE, echo=FALSE,  width=12, height=6.8>>=
plot(ct)
@
\end{minipage}
\caption{\label{fig:example_rpart_ctree} Trees constructed by \code{rpart} (upper panel) and \code{ctree} (lower panel). The target variable is the customers choice of buying the book. The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase; \code{freq} number of books purchased at the Bookbinder's Book Club; \code{gender} the customers gender.
}
\end{figure}
%
With the objective of building a smaller, but at least equally accurate tree, \code{evtree} is constrained to a maximum tree depth of $2$. The resulting tree is displayed in Figure~\ref{fig:example_evtree}. Alternatively, the tree can be examined by the textual output provided by the \code{print()} function.
%
<<evtree-plot>>=
ev <- evtree(choice ~ ., data = BBBClub, minbucket = 26, maxdepth = 2)
plot(ev)
print(ev)
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\begin{flushleft} \code{evtree} \end{flushleft}
\centering
<<iris-plot, fig=TRUE, echo=FALSE, width=12, height=7>>=
plot(ev)
@
\caption{\label{fig:example_evtree} Tree constructed by \code{evtree}. The target variable is the customers choice of buying the book. The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase.}
\end{figure}
%
While the \code{rpart} model has $5$ terminal nodes and a tree depth of $4$, the tree constructed by \code{ctree} has $7$ terminal nodes and a tree depth of $3$. With a tree depth of $2$ and $4$ terminal nodes the \code{evtree} model is much simpler. 
%
<<evtree-performance>>=
## accuracy: misclassification rate
mc <- function(obj) 1 - mean(predict(obj) == BBBClub$choice)
c("evtree" = mc(ev), "rpart" = mc(rp), "ctree" = mc(ct))

## complexity: number of terminal nodes
c("evtree" = width(ev), "rpart" = width(rp), "ctree" = width(ct))
@
The accuracy of the three models are similar. With a misclassification rate of $24.3\%$, the \code{evtree} model is slightly worse than the \code{rpart} model ($23.8\%$) and better than the \code{ctree} model ($24.9\%$). 
%
<<evtree-structure>>=
## compare structure of predictions
ftable(tab <- table(evtree = predict(ev), rpart  = predict(rp),
  ctree  = predict(ct), observed = BBBClub$choice))

## compare customer predictions only (absolute, proportion correct)
sapply(c("evtree", "rpart", "ctree"), function(nam) {
  mt <- margin.table(tab, c(match(nam, names(dimnames(tab))), 4))
  c(abs = as.vector(rowSums(mt))[2],
    rel = round(100 * prop.table(mt, 1)[2, 2], digits = 3))
})
@
%
Although the constructed trees are considerably different, the structure of the predictions are relatively similar. In this case, \code{evtree} classifies less customers ($186$) as buyers as \code{rpart} ($216$) and \code{ctree} ($278$). Of the customers classified as buyers, $72.6\%$ (\code{evtree}), $70.8\%$ (\code{rpart}) and $63.7\%$ (\code{ctree}) actually bought the book. In a final step \code{rpart} and \code{ctree} models are fitted with the same constrained on the tree's depth (\code{maxdepth = 2}) as the previously fitted \code{evtree} model.
%
<<evtree-performance>>=
rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 26, 
maxdepth = 2))
ct <- ctree(choice ~ ., data = BBBClub, mincrit = 0.99, minbucket = 26, 
maxdepth = 2)
## accuracy: misclassification rate
mc <- function(obj) 1 - mean(predict(obj) == BBBClub$choice)
c("evtree" = mc(ev), "rpart" = mc(rp), "ctree" = mc(ct))
## complexity: number of terminal nodes
c("evtree" = width(ev), "rpart" = width(rp), "ctree" = width(ct))
@
%
With a misclassification rate of $24.3\%$, the \code{evtree} model is more accurate than the \code{rpart} model ($26.2\%$) and the \code{ctree} model ($25.5\%$). Due to its unbalanced structure, the \code{rpart} model has only $3$ terminal nodes. Hence, it does not use the maximum tree depth efficiently. In contrast, a global optimization algorithm like \code{evtree} can be used to search the best tree under given constraints.

>>>>

This section continues with a sample session where \code{evtree} is used, in conjunction with \code{rpart} and \code{ctree}, to analyze the \textit{Bookbinder's Book Club} data. In this session, all trees are constrained to have a minimum number of $10$ observations per terminal node. To avoid excessively large trees \code{ctree} is called with $\alpha = 1\%$. To provide uniform visualizations and predictions of the fitted models the \code{as.party()} function is used to convert \pkg{rpart} objects into objects of class \pkg{party}. The sample session starts with the construction and visualization of trees using the \code{rpart} and \code{ctree} methods.
%
<<rpart-ctree>>=
data("BBBClub", package = "evtree")
library("rpart")
set.seed(1090)
rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10))
ct <- ctree(choice ~ ., data = BBBClub, mincrit = 0.99, minbucket = 10)
plot(rp)
plot(ct)
@
%
The resulting trees are displayed in Figure~\ref{fig:example_rpart_ctree}.
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\centering
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{rpart} \end{flushleft}
\centering
<<rp-plot, fig=TRUE, echo=FALSE, width=12, height=6.8>>=
plot(rp)
@
\end{minipage}
\begin{minipage}{1\textwidth}
\begin{flushleft} \code{ctree} \end{flushleft}
\centering
<<ct-plot, fig=TRUE, echo=FALSE,  width=12, height=6.8>>=
plot(ct)
@
\end{minipage}
\caption{\label{fig:example_rpart_ctree} Trees constructed by \code{rpart} (upper panel) and \code{ctree} (lower panel). The target variable is the customers choice of buying the book. The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase; \code{freq} number of books purchased at the Bookbinder's Book Club; \code{gender} the customers gender.
}
\end{figure}
%
With the objective of building a smaller, but at least equally accurate tree, \code{evtree} is constrained to a maximum tree depth of $2$. The resulting tree is displayed in Figure~\ref{fig:example_evtree}. Alternatively, the tree can be examined by the textual output provided by the \code{print()} function.
%
<<evtree-plot>>=
ev <- evtree(choice ~ ., data = BBBClub, minbucket = 10, maxdepth = 2)
plot(ev)
print(ev)
@
%
\setkeys{Gin}{width=1\textwidth}
\begin{figure}[t!]
\begin{flushleft} \code{evtree} \end{flushleft}
\centering
<<iris-plot, fig=TRUE, echo=FALSE, width=12, height=7>>=
plot(ev)
@
\caption{\label{fig:example_evtree} Tree constructed by \code{evtree}. The target variable is the customers choice of buying the book. The variables used for splitting are: \code{art} number of art books purchased; \code{first} number of months since the first purchase.}
\end{figure}
%
While the \code{rpart} model has $5$ terminal nodes and a tree depth of $4$, the tree constructed by \code{ctree} has $7$ terminal nodes and a tree depth of $3$. With a tree depth of $2$ and $4$ terminal nodes the \code{evtree} model is much simpler. 
%
<<evtree-performance>>=
mc <- function(obj) 1 - mean(predict(obj) == BBBClub$choice)
c("evtree" = mc(ev), "rpart" = mc(rp), "ctree" = mc(ct))
@
The accuracy of the three models are similar. With a misclassification rate of $24.3\%$, the \code{evtree} model is slightly worse than the \code{rpart} model ($23.8\%$) and better than the \code{ctree} model ($24.8\%$). 
%
<<evtree-structure>>=
ftable(tab <- table(evtree = predict(ev), rpart  = predict(rp),
  ctree  = predict(ct), observed = BBBClub$choice))

sapply(c("evtree", "rpart", "ctree"), function(nam) {
  mt <- margin.table(tab, c(match(nam, names(dimnames(tab))), 4))
  c(abs = as.vector(rowSums(mt))[2],
    rel = round(100 * prop.table(mt, 1)[2, 2], digits = 3))
})
@
%
Although the constructed trees are considerably different, the structure of the predictions are relatively similar. In this case, \code{evtree} classifies less customers ($186$) as buyers as \code{rpart} ($216$) and \code{ctree} ($238$). Of the customers classified as buyers, $72.6\%$ (\code{evtree}), $70.8\%$ (\code{rpart}) and $66.4\%$ (\code{ctree}) actually bought the book. In a final step \code{rpart} and \code{ctree} models are fitted with the same constrained on the tree's depth (\code{maxdepth = 2}) as the previously fitted \code{evtree} model.
%
<<evtree-performance>>=
rp <- as.party(rpart(choice ~ ., data = BBBClub, minbucket = 10, 
maxdepth = 2))
ct <- ctree(choice ~ ., data = BBBClub, mincrit = 0.99, minbucket = 10, 
maxdepth = 2)

mc <- function(obj) 1 - mean(predict(obj) == BBBClub$choice)
c("evtree" = mc(ev), "rpart" = mc(rp), "ctree" = mc(ct))

c("evtree" = width(ev), "rpart" = width(rp), "ctree" = width(ct))
@
%
With a misclassification rate of $24.3\%$, the \code{evtree} model is more accurate than the \code{rpart} model ($26.2\%$) and the \code{ctree} model ($25.5\%$). Due to its unbalanced structure, the \code{rpart} model has only $3$ terminal nodes. Hence, it does not use the maximum tree depth efficiently. In contrast, a global optimization algorithm like \code{evtree} can be used to search the best tree under given constraints.
